Is essentiality of genes conserved in Enterobacteriaceae?

### 2017-03-29 documents/manuscript.tex
Made a table of unannotated genes that we had annotated using prokka, added duplicated genes to heatmap

%\begin{figure*}
%\begin{tabular}{c c}
%\includegraphics[scale=0.4, page=1]{insertion-position-bias.pdf}&
%\includegraphics[scale=0.4, page=2]{insertion-position-bias.pdf}\\
%\includegraphics[scale=0.4, page=3]{insertion-position-bias.pdf}&
%\includegraphics[scale=0.4, page=4]{insertion-position-bias.pdf}
%\end{tabular}
%\caption{The plots show the average insertion index in the percentiles of all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right). The genes are divided into 3 segments: 5\% of
%the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
%\label{fig:insertion-position-bias}
%\end{figure*}
\subsection{Most essential genes are ubiquitously essential in Enterobacteriaceae}
Previous studies of gene essentiality in Enterobacteriaceae family have compared essential genes in different genomes in this family and studied the sets of core essential genes and accessory essential genes \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}. Core essential genes are responsible for essential processes such as cell division, DNA replication, transcription and translation and some important pathways like peptidoglycan and fatty acid biosynthesis \cite{barquist_comparison_2013}. Accessory essential genes differ in genomes due to different reasons such as niche adaptation, gene duplication and the existence of alternative pathways \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}. Another group of accessory essential genes are phage repressors \cite{barquist_comparison_2013}. Even though these genes are not essential for the growth of a cell, once phages are introduced to a cell, they become essential as long as the phage remains in the cell.

In this study, we have compared the genes in 16 bacteria from Enterobacteriaceae family. For this we needed to study core and accessory sets of genes. Moreover, in the presence of two redundant variations of one gene, if we knock out one copy using TraDIS, the other copy compensates and the organism can still survive \cite{dean_pervasive_2008}. This leads to a different essentiality inference using TraDIS. Therefore, in addition to core and accessory genes, we have studied duplicate genes. To study whether each gene in the 16 organisms is core, we used Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to iteratively search for homologous proteins in our dataset and cluster them. We divided the clusters of homologous genes into three groups based on their conservation. Genus specific class contains genes that are present only in one genus, the genes in single copy class are present in more than one genus and more than 70\% of them are not duplicated (core genes), and  the genes in multi-copy class are present in more than one genus and more than 30\% of them are duplicated (duplicate genes).

@@

\subsection{Phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}
%When a mutation is introduced to a gene, the fixation rate depends on how tolerable this mutation is \cite{wilson_biochemical_1977}. As mutations in essential genes are less likely to be tolerable, less mutations should be fixed in these genes, and therefore essential genes should evolve more slowly than non-essential genes. This is known as knockout rate hypothesis \cite{krylov_gene_2003}. Different studies have verified this hypothesis in prokaryotes \cite{jordan_essential_2002, gerdes_experimental_2003, rocha_analysis_2004, silander_constancy_2009}.

Here, we have tested this hypothesis by comparing the number of core essential genes and core genes in the phylogenetic tree as we go up to the root. We have used Fitch's algorithm \cite{fitch_toward_1971} with a binary alphabet on both essentiality (0 for non-essential and 1 for essential) and conservation (1 for the presence and 0 for the absence of genes) to define if a gene is core essential or core at each level in the phylogenetic tree. If we get 1 at the top of the tree while studying essentiality of a gene using Fitch algorithm, it means that the gene is core essential and otherwise, the gene is non-essential. The same applies to the study of core genes. The phylogenetic tree has been annotated with the number of core essential genes (red) and the number of core genes (blue) at each level in Figure~\ref{fig:tree}. We then plotted the ratios at each level in the phylogenetic tree in Figure~\ref{fig:fitch} and connected the medians in each level. The connecting line shows that the ratio between core essential genes and core genes rises as we go higher in the phylogenetic tree which means essential genes are more likely to be conserved in genomes compared to non-essential genes.

\begin{figure}
\captionsetup[subfigure]{justification=centering}
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.12]{phylosift-aa-raxmlbootstrap-annotated.pdf}
  \caption{}
  \label{fig:tree}
\end{subfigure}
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[scale=0.4]{fitch.pdf}
  \caption{}
  \label{fig:fitch}
\end{subfigure}
\caption{(a) The species tree for all genomes in this study. Numbers in red show the number of core essential genes at each level and numbers in blue show the number of core genes at each level. (b) The ratio between core essential genes and core genes at each level in the species tree. The dots in the strain level show the ratios for all 16 bacteria; the dots in the species level show the ratios for Enterobacter, Klebsiella, Citrobacter, Salmonella, and Escherichia; the subfam. level shows the ratios for the common ancestor of Enterobacter and klebsiella, the common ancestor of Citrobacter and Salmonella, and the common ancestor of Citrobacter, Salmonella, and Escherichia; and finally the dot in the family level shows the ratio for the root. The line connects the medians in each level.}
\label{fig:essentiality-phylogeny}
\end{figure}

We have studied the essentiality status of genes involved in important biological processes in Table~\ref{tab:biological-processes}. These processes include cell division, DNA replication, transcription, translation, and important metabolic pathways such as peptidoglycan and fatty acid biosynthesis. If a gene was not core essential using Fitch algorithm, we looked at its essentiality in every genome: if it was essential in some of the genomes we classified it as accessory essential and if it was not essential in any genome we classified it as non-essential. FtsA and zipA genes are essential for cell division. FtsA is essential in all the genomes in our study, however zipA is essential everywhere except for Salmonella typhimurium A130 in which this gene is near essential. We investigated if this gene has a homolog in Salmonella typhimurium A130 and found no homologs. It has been shown that 21 mutations in ftsA gene can make it independent of zipA gene so that the zipA gene is no longer essential \cite{geissler_gain--function_2003, pichoff_ftsa_2012}. However, ftsA gene does not have any of these mutations in Salmonella typhimurium A130, so there might be another factor that is causing zipA not to be essential. \textcolor{red}{ADD MORE EXAMPLES HERE}

So far we have shown that essential genes are less likely to be lost during evolution. The other question is if essential genes tend to keep their essentiality in different genomes. Bergmiller et al. \cite{bergmiller_patterns_2012} have shown that only essential genes whose functions cannot be taken over by other genes keep their essentiality and others are possible to be lost or non-essential in other genomes. The genes that can replace other genes are not necessarily homologous to them, but they have similar functions.

We have compared the number of genes that are shared between our bacterial genomes (\ref{fig:upsetr-conservation}) and the number of genes essential in them (\ref{fig:upsetr-essentiality}) and used UpSetR package \cite{conway_upsetr:_2016} to visualise the results in Fig.\@ \ref{fig:upsetr}. As shown in the figures, among 2162 genes that are shared between all the bacteria under study, only 135 are essential everywhere. We looked at subsets of the genes that were shared in every combination of our bacterial strains to see whether they show a trend similar to the phylogenetic tree. The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show such a trend. The reason is that many of the essential genes that are not shared in all genomes are probably compensable by other genes and so the essentiality does not look tree-like.

\begin{landscape}
\begin{table}
\small
\caption{The essentiality status of genes involved in important biological processes}
\label{tab:biological-processes}
\begin{tabular}[t]{l l l l l}
\hline
\textbf{Biological process} & \textbf{Subprocess} & \textbf{Core essential} & \textbf{Accessory essential} & \textbf{Non-essential}\\
\hline
Cell Division & & \begin{tabular}[t]{@{}l@{}}ftsAHLQWYZ, minE,\\ mukB, zipA \end{tabular}& ftsKNX, minD & \begin{tabular}[t]{@{}l@{}}CedA, ftsJ, minC,\\ sdiA, sulA\end{tabular}\\
DNA replication & Polymerases I, II, III & dnaENQX, holABD, polA & holC & holE, polB\\	
& Supercoiling & gyrAB, parCE & & \\
& Primosome-associated & dnaBCGT, priA, ssb & priB, rep & priC\\
Transcription & RNA polymerase & rpoABC &  & \\
& \begin{tabular}[t]{@{}l@{}}Sigma, elongation, anti- \\ and termination factors\end{tabular} & nusABG, rho, rpoDH & rpoEN & rpoS\\	
Translation & tRNA-synthetases & \begin{tabular}[t]{@{}l@{}}alaS, argS, asnS, aspS, cysS, \\ glnS, gltX, glyQS, hisS, ileS, \\ leuS, lysS, metG, pheST, \\ proS, serS, thrS, tyrS, valS \end{tabular} & trpS	 & \\	
& Ribosome components & \begin{tabular}[t]{@{}l@{}}rplBCDEFJKLMNOPQRSTUV,\\ rplWXY, rpmABCD,\\ rpsABCDEFGHJKLMNPQRSTU\end{tabular} & rplA, rpmEGHI, rpsIO & rplI, rpmF\\
& \begin{tabular}[t]{@{}l@{}}Initiation, elongation \\ and peptide chain \\ release factors\end{tabular} & fusA, infABC, prfAB, tsf & efp & prfC, selB, tufAB \\
Biosynthetic pathways & & & & \\
Peptidoglycan & & MraY, murABCDEFGI & & ddlAB\\	
Fatty acids & & accABCD, fabABDGHIZ & & \\
\hline
\end{tabular}
\end{table}
\end{landscape}

>>>>

%\begin{figure*}
%\begin{tabular}{c c}
%\includegraphics[scale=0.4, page=1]{insertion-position-bias.pdf}&
%\includegraphics[scale=0.4, page=2]{insertion-position-bias.pdf}\\
%\includegraphics[scale=0.4, page=3]{insertion-position-bias.pdf}&
%\includegraphics[scale=0.4, page=4]{insertion-position-bias.pdf}
%\end{tabular}
%\caption{The plots show the average insertion index in the percentiles of all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right). The genes are divided into 3 segments: 5\% of
%the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
%\label{fig:insertion-position-bias}
%\end{figure*}
\subsection{Most essential genes are ubiquitously essential in Enterobacteriaceae}
Previous studies of gene essentiality in Enterobacteriaceae family have compared essential genes in different genomes in this family and studied the sets of core essential genes and accessory essential genes \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}. Core essential genes are responsible for essential processes such as cell division, DNA replication, transcription and translation and some important pathways like peptidoglycan and fatty acid biosynthesis \cite{barquist_comparison_2013}. Accessory essential genes differ in genomes due to different reasons such as niche adaptation, functional redundancy and the existence of alternative pathways \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013, bergmiller_patterns_2012}. Another group of accessory essential genes are phage repressors \cite{barquist_comparison_2013}. Even though these genes are not essential for the growth of a cell, once phages are introduced to a cell, they become essential as long as the phage remains in the cell.

In this study, we have compared the genes in 16 bacteria from Enterobacteriaceae family. For this we needed to study core and accessory sets of genes. Moreover, in the presence of two redundant variations of one gene, if we knock out one copy using TraDIS, the other copy compensates and the organism can still survive \cite{bergmiller_patterns_2012,dean_pervasive_2008}. This leads to a different essentiality inference using TraDIS. Therefore, in addition to core and accessory genes, we have studied duplicate genes. To study whether each gene in the 16 organisms is core, we used Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to iteratively search for homologous proteins in our dataset and cluster them. We divided the clusters of homologous genes into three groups based on their conservation. Genus specific class contains genes that are present only in one genus, the genes in single copy class are present in more than one genus and more than 70\% of them are not duplicated (core genes), and  the genes in multi-copy class are present in more than one genus and more than 30\% of them are duplicated (duplicate genes).

@@

%\subsection{Phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}

%When a mutation is introduced to a gene, the fixation rate depends on how tolerable this mutation is \cite{wilson_biochemical_1977}. As mutations in essential genes are less likely to be tolerable, less mutations should be fixed in these genes, and therefore essential genes should evolve more slowly than non-essential genes. This is known as knockout rate hypothesis \cite{krylov_gene_2003}. Different studies have verified this hypothesis in prokaryotes \cite{jordan_essential_2002, gerdes_experimental_2003, rocha_analysis_2004, silander_constancy_2009}.

We have shown that many of the essential genes are conserved. But are essential genes more likely to be conserved? To answer this question, we have compared the number of ancestrally essential genes and ancestrally present genes in the phylogenetic tree as we go up to the root. We have used Fitch's algorithm \cite{fitch_toward_1971} with a binary alphabet on both essentiality (0 for non-essential and 1 for essential) and conservation (1 for the presence and 0 for the absence of genes) to define if a gene is ancestral essentially or present at each level in the phylogenetic tree. If we get 1 at the top of the tree while studying essentiality of a gene using Fitch's algorithm, it means that the gene is ancestrally essential and otherwise, the gene is ancestrally non-essential. The same applies to the study of ancestrally present genes. The phylogenetic tree has been annotated with the number of ancestrally essential genes (red) and the number of ancestral genes (blue) at each level in Figure~\ref{fig:tree}. We then plotted the ratios at each level in the phylogenetic tree in Figure~\ref{fig:fitch} and connected the medians in each level. The connecting line shows that the ratio between ancestrally essential genes and ancestral genes rises as we go higher in the phylogenetic tree which means essential genes are more likely to be conserved in genomes compared to non-essential genes.

\begin{figure}
\captionsetup[subfigure]{justification=centering}
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.12]{phylosift-aa-raxmlbootstrap-annotated.pdf}
  \caption{}
  \label{fig:tree}
\end{subfigure}
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[scale=0.4]{fitch.pdf}
  \caption{}
  \label{fig:fitch}
\end{subfigure}
\caption{(a) The species tree for all genomes in this study. Numbers in red show the number of ancestrally essential genes at each level and numbers in blue show the number of ancestral genes at each level. (b) The ratio between ancestrally essential genes and ancestral genes at each level in the species tree. The dots in the strain level show the ratios for all 16 bacteria; the dots in the species level show the ratios for Enterobacter, Klebsiella, Citrobacter, Salmonella, and Escherichia; the subfam. level shows the ratios for the common ancestor of Enterobacter and klebsiella, the common ancestor of Citrobacter and Salmonella, and the common ancestor of Citrobacter, Salmonella, and Escherichia; and finally the dot in the family level shows the ratio for the root. The line connects the medians in each level.}
\label{fig:essentiality-phylogeny}
\end{figure}

We have studied the essentiality status of genes involved in important biological processes in Table~\ref{tab:biological-processes}. These processes include cell division, DNA replication, transcription, translation, and important metabolic pathways such as peptidoglycan and fatty acid biosynthesis. If a gene was not ancestrally essential using Fitch's algorithm, we looked at its essentiality in every genome: if it was essential in some of the genomes we classified it as ambiguous and if it was not essential in any genome we classified it as ancestrally non-essential.
%FtsA and zipA genes are essential for cell division. FtsA is essential in all the genomes in our study, however zipA is essential everywhere except for Salmonella typhimurium A130 in which this gene is near essential. We investigated if this gene has a homolog in Salmonella typhimurium A130 and found no homologs. It has been shown that 21 mutations in ftsA gene can make it independent of zipA gene so that the zipA gene is no longer essential \cite{geissler_gain--function_2003, pichoff_ftsa_2012}. However, ftsA gene does not have any of these mutations in Salmonella typhimurium A130, so there might be another factor that is causing zipA not to be essential. \textcolor{red}{ADD MORE EXAMPLES HERE}

\begin{landscape}
\begin{table}
\small
\caption{The essentiality status of genes involved in important biological processes}
\label{tab:biological-processes}
\begin{tabular}[t]{l l l l l}
\hline
\textbf{Biological process} & \textbf{Subprocess} & \textbf{Ancestrally essential} & \textbf{Ambiguous} & \textbf{Ancestrally non-essential}\\
\hline
Cell Division & & \begin{tabular}[t]{@{}l@{}}ftsAHLQWYZ, minE,\\ mukB, zipA \end{tabular}& ftsKNX, minD & \begin{tabular}[t]{@{}l@{}}CedA, ftsJ, minC,\\ sdiA, sulA\end{tabular}\\
DNA replication & Polymerases I, II, III & dnaENQX, holABD, polA & holC & holE, polB\\	
& Supercoiling & gyrAB, parCE & & \\
& Primosome-associated & dnaBCGT, priA, ssb & priB, rep & priC\\
Transcription & RNA polymerase & rpoABC &  & \\
& \begin{tabular}[t]{@{}l@{}}Sigma, elongation, anti- \\ and termination factors\end{tabular} & nusABG, rho, rpoDH & rpoEN & rpoS\\	
Translation & tRNA-synthetases & \begin{tabular}[t]{@{}l@{}}alaS, argS, asnS, aspS, cysS, \\ glnS, gltX, glyQS, hisS, ileS, \\ leuS, lysS, metG, pheST, \\ proS, serS, thrS, tyrS, valS \end{tabular} & trpS	 & \\	
& Ribosome components & \begin{tabular}[t]{@{}l@{}}rplBCDEFJKLMNOPQRSTUV,\\ rplWXY, rpmABCD,\\ rpsABCDEFGHJKLMNPQRSTU\end{tabular} & \begin{tabular}[t]{@{}l@{}}rplA, rpmEGHI,\\ rpsIO\end{tabular} & rplI, rpmF\\
& \begin{tabular}[t]{@{}l@{}}Initiation, elongation \\ and peptide chain \\ release factors\end{tabular} & fusA, infABC, prfAB, tsf & efp & prfC, selB, tufAB \\
Biosynthetic pathways & & & & \\
Peptidoglycan & & MraY, murABCDEFGI & & ddlAB\\	
Fatty acids & & accABCD, fabABDGHIZ & & \\
\hline
\end{tabular}
\end{table}
\end{landscape}

%So far we have shown that essential genes are less likely to be lost during evolution. The other question is if essential genes tend to keep their essentiality in different genomes. Bergmiller et al. \cite{bergmiller_patterns_2012} have shown that only essential genes whose functions cannot be taken over by other genes keep their essentiality and others are possible to be lost or non-essential in other genomes. The genes that can replace other genes are not necessarily homologous to them, but they have similar functions.
Genes that are essential in one organism can be non-essential or absent in other organisms. Bergmiller et al. \cite{bergmiller_patterns_2012} have studied 26 genes that are essential in E. \textit{coli}. Some of these genes are essential in other bacteria and some are non-essential or not present in other bacteria. In 10 cases, they were able to find genes that compensate for these essential genes if they are overexpressed and called these genes high copy suppressors. High copy suppressors are not necessarily similar to their counterpart essential genes in structure or sequence, but they have similar functions. They have shown that genes that are not present or essential everywhere are more likely to be compensated for by overexpression of another gene. The other reason for genes to lose their essentiality in some strains is changes in their physiology or environment \cite{barquist_comparison_2013}.

We have compared the number of genes that are shared between our bacterial genomes (\ref{fig:upsetr-conservation}) and the number of genes essential in them (\ref{fig:upsetr-essentiality}) and used UpSetR package \cite{conway_upsetr:_2016} to visualise the results in Fig.\@ \ref{fig:upsetr}. As shown in the figures, among 2162 genes that are shared between all the bacteria under study, only 135 are essential everywhere and many of the essential genes are not conserved or essential everywhere probably because of their compensability, physiological, or environmental changes. %We looked at subsets of the genes that were shared in every combination of our bacterial strains to see whether they show a trend similar to the phylogenetic tree. The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show such a trend. The reason is that many of the essential genes that are not shared in all genomes are probably compensable by other genes and so the essentiality does not look tree-like.
